Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-04-14 Paper/Paper.tex

The first dependent variable I consider is the decision for a physician to retire, or more broadly to stop practicing in a clinical setting altogether. Retirement is constructed as follows: 
$$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$. Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future. I impose that no retirement happens in 2016 since dropping out of the data for one year does not definitively show retirement. Alternatively, I could also define retirement using future number of patients. I use claim count for a more conservative measure of retirement, but using patient count yields identical results.

@@

There are several assumptions necessary to identify $ATT(g,t)$. First, I assume that treatment is not reversed once it occurs. That is, once a physician is exposed to an EHR, they cannot be un-exposed. This assumption is supported by both the institutional details of hospital EHR implementation and the definition of exposure. An EHR is a costly technology and requires a significant amount of collaboration to implement. A hospital does not have incentive to un-implement an EHR once it is implemented. They may add features or switch vendors, but do not reverse EHR use. Further, since we think of treatment at the EHR level as exposure to the technology, this assumption is automatically satisfied when thought of as whether the physician has ever used an EHR. Once they are exposed to the EHR for the first time, that experience cannot be forgotten. Thus, physicians remain treated in all future years after exposure.

Second, I assume physicians do not anticipate EHR exposure prior to occurrence. Since the technology has different vendors and varies in capabilities, anticipation may be a concern if physicians learn that the system will be implemented and then do not actually use it until years later. Generally, even a complex EHR system can be completely set up within a year, and most systems take 6 to 9 months (\cite{uzialko_2021}). Therefore, I proceed in the main specification assuming no anticipation. However, I explore and present results for a one year anticipation period in Appendix ().

@@

It is not uncommon in developed countries for workers to plan for formal retirement well in advance. An exogenous shock to a worker is not expected to change retirement age because of the amount of wealth and planning necessary to formally leave the labor force. However, physicians typically make this decision differently than workers in other industries. Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}). When the time to retire comes, many physicians hesitate to abandon patients they have seen over the course of their career. The decision to delay retirement is not financial, but altruistic. Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force. Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role. There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, or hospital management. While it is unlikely that physicians under 55 are formally retiring, either old or young physicians could be career switching. The term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no claim about what the physician does afterwards. A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.

An exogenous shock such as EHR implementation in place of work may cause an earlier exit than would have occurred without the shock. Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}). The aim of this section is to test that hypothesis.

In Figure \ref{fig:retirefirst}, I present aggregated group time treatment effects of being exposed to an EHR in any hospital on the likelihood of retirement for the full sample of physicians, as well as split samples for those at least 60 and less than 60 years of age. The top panel suggests that for any age physician, being exposed to an EHR leads to a .002 ppt increase in the likelihood of retiring in both the first and second year after after implementation. While this result is numerically small, it is statistically and economically meaningful relative to the proportion of physicians who actually retire in the sample, .02. Thus, relative to the mean, this effect represents a 10\% increase in the likelihood of retirement due to EHR exposure. It would be alarming to see numerically large effects on this outcome since retirement is a massive and one-time decision that is affected by many exogenous factors. While the presented estimates are noisy, the rarity of retirement in the sample combined with the observed precision in the estimates indicates that retirement did occur as a result of EHR implementation.

Now, I examine how these result differ when considering physicians in different age brackets: pre-retirement and typical retirement age. These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the result in the first year after EHR exposure. A physician who is at least 60 years old is .005 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean. In comparison, physicians less than 60 are far less likely, if at all, to respond to EHR exposure in the first year after EHR exposure. Physicians in the young category have a slightly increased likelihood of retiring in the second year after exposure relative to the first year. While I do not have data to test this, if young physicians are more likely to switch careers after EHR exposure then it may take one additional year to prepare for that decision relative to retirement.

@@

I first discuss the effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}. The top panel is for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .015 percentage points, a 2.6\% increase relative to the mean. Splitting the sample by age reveals that this result is driven by the younger sample of physicians. There is no evidence that older physicians are shifting towards offices. This may be due to senior physicians already being established in some capacity in offices prior to EHR exposure, but is also supported by intuition that older workers are more "stuck" to their particular work environment.

If senior physicians are already established in offices in some capacity, they may be more affected on an intensive margin, the fraction of patients seen in an office relative to hospital or other settings. The estimates for this outcome are shown in Figure \ref{fig:officesecond}, where I find limited evidence that EHR exposure affects the fraction of patients seen in office settings. Of course, this could mean multiple things. If total number of patients remains constant, then this implies that the absolute number of patients seen in offices also is unchanged. However, if the total number of patients seen (which depends on efficiency in hospitals also) changes, then the absolute number of patients seen in offices necessarily changes in a similar direction. I revisit the implications of this result in the next subsection where I examine the total number of patients as an outcome variable.

A much more general way to examine  physician switching behavior is to investigate whether the zip codes in which the physicians work is changing. Again, I exclude physicians who ever retire. The outcome variable is equal to 1 if the list of primary zip codes (up to twelve of them) changes. This variable is limited in the way it captures switching behavior because multiple scenarios are flagged as changing zip codes: leaving a particular zip code, adding one new zip code but leaving the rest unchanged, or completely changing zip codes. While this is a vague description of behavior, it provides a different insight into changes occurring because of EHR implementation.

@@

\subsection{Patient Count}

An important, and heavily debated, aspect of the implementation of EHRs is whether or not they increase productivity in users. I consider productivity and efficiency as interchangeable, the interest here is the number of patients a physician can see in a given amount of time. Importantly, my measure of efficiency does not necessarily indicate quality. Nevertheless, health care in the U.S. can be wasteful relative to the resources used to produce it. I make no claims about whether the number of patients seen is optimal, I simply explore whether the number of patients seen changes as a result of EHR exposure since a primary intention of the technology is to improve efficiency.

In this case, efficiency has two key determinants: the choice of physicians to learn and utilize the technology, and the realized quality/innovation of electronic health records themselves. Regarding the first determinant, I will assume that when a physician continues to work in an electronic record utilizing-hospital, they utilize the electronic health record system fully. That is, there are no physicians who stay actively working in a hospital but choose not to partake in the electronic health record used by the hospital. To understand the second determinant, the value added of electronic health records themselves, the sample of physicians is limited to those who have a positive number of patients in hospitals even after EHR exposure.

@@

The way that physicians bill for claims is of extreme interest to researchers because incentives to report certain claims do not always line up with optimizing both outcomes and efficiency of care. EHRs have the potential to change claim incentives if the administrative costs of billing decrease or if EHRs encourage billing for more items than otherwise would have been recorded.

@@

The total number of registered data assistants may not equal the true number of people doing a data assistant job in hospitals. However, under the assumption that the number of registered data assistants is proportional to the number of actual data assistants, the claim of no correlation still holds.

@@

>>>>

The first dependent variable I consider is the decision for a physician to retire, or more broadly to stop practicing in a clinical setting altogether. Retirement is constructed as follows: 
$$\text{retire}_{it}=\mathbbm{1}\{FC_{it}=0 \text{ and retire}_{i,t-1}=0\}, $$
where $FC_{it}=\sum\limits_{j=t+1}^T\text{(claim count)}_{ij}$ is all future claims for physician $i$ in year $t$. Retire is therefore an indicator variable set to one in the first year that a physician has no claim counts at any point in the future. I impose that no retirement happens in 2016 since dropping out of the data for one year does not definitively show retirement. \footnote{Alternatively, I could also define retirement using future number of patients. I use claim count for a more conservative measure of retirement, but using patient count yields identical results.}

@@

There are several assumptions necessary to identify $ATT(g,t)$. First, I assume that treatment is not reversed once it occurs. That is, once a physician is exposed to an EHR, they cannot be un-exposed. This assumption is supported by both the institutional details of hospital EHR implementation and the definition of exposure. An EHR is a costly technology and requires a significant amount of collaboration to implement. A hospital does not have incentive to un-implement an EHR once it is implemented. They may add features or switch vendors, but do not reverse EHR use. This institutional detail is supported by Figure \ref{fig:hosp_treat}, which shows EHR status of a random sample of hospitals over time. Further, since we think of treatment at the EHR level as exposure to the technology, this assumption is automatically satisfied when thought of as whether the physician has ever used an EHR. Once they are exposed to the EHR for the first time, that experience cannot be forgotten. Thus, physicians remain treated in all future years after exposure.

Second, I assume physicians do not anticipate EHR exposure prior to occurrence. Since the technology has different vendors and varies in capabilities, anticipation may be a concern if physicians learn that the system will be implemented and then do not actually use it until a future period. Generally, even a complex EHR system can be completely set up within a year, and most systems take 6 to 9 months (\cite{uzialko_2021}). Therefore, I proceed in the main specification assuming no anticipation. However, I explore and present results for a one year anticipation period in Appendix ().

@@

Workers in developed countries tend to plan for formal retirement well in advance. An exogenous shock to a worker is not expected to change retirement age because of the amount of wealth and planning necessary to formally leave the labor force. However, physicians typically make this decision differently than workers in other industries. Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}). When the time to retire comes, many physicians hesitate to abandon patients they have seen over the course of their career. The decision to delay retirement is not financial, but altruistic. Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force. Formal retirement (leaving the labor force altogether) is not the only way for physicians to transition out of a clinical role. There are opportunities for physicians of any age to switch careers towards more administrative roles such as teaching, consulting, or hospital management. While it is unlikely that physicians under 55 are formally retiring, physicians of any age could be career switching. The term "retirement" for the purpose of this paper indicates no longer seeing patients, and makes no assertion about what the physician does afterwards. A physician is said to retire in a given year only if it is the first year in which the physician has no future Medicare claims.

An exogenous shock to the physician such as EHR implementation in place of work may cause an earlier exit than would have occurred without the shock. Media articles focusing on interviews with particular physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}). The aim of this section is to test that hypothesis.

In Figure \ref{fig:retirefirst}, I present aggregated group time treatment effects of being exposed to an EHR in any hospital on the likelihood of retirement for the full sample of physicians, as well as split samples for those at least 60 and less than 60 years of age. The top panel suggests that for any age physician, being exposed to an EHR leads to a .002 ppt increase in the likelihood of retiring in both the first and second year after after implementation. While this result is numerically small, it is statistically and economically meaningful relative to the proportion of physicians who actually retire in the sample, .02. Thus, relative to the mean, this effect represents a 10\% increase in the likelihood of retirement due to EHR exposure. In this scenario, it would be alarming to see a numerically large effect on this outcome since retirement is a massive and one-time decision that is affected by many exogenous factors. While the presented estimates are noisy, the rarity of retirement in the sample combined with the observed precision in the estimates indicates that a positive effect on the likelihood of retirement did occur as a result of EHR implementation.

Now, I examine how these result differ when considering physicians in different age brackets: pre-retirement and typical retirement age. These results are shown in the bottom panels of Figure \ref{fig:retirefirst}, and suggest that senior physicians are driving the result in the first year after EHR exposure. A physician who is at least 60 years old is .005 ppts more likely to retire after being exposed to an EHR, a 25\% increase relative to the mean. In comparison, physicians less than 60 are far less likely, if at all, to respond to EHR exposure in the first year after EHR exposure. Physicians in the young category have a larger effect on the likelihood of retiring in the second year after exposure relative to the first year. While I do not have data to test this, if young physicians are more likely to switch careers after EHR exposure then it may take one additional year to prepare for that decision relative to retirement.

@@

I first discuss the effect of EHR exposure on the probability of working in an office setting, shown in Figure \ref{fig:officefirst}. The top panel shows aggregated group time effects for physicians of any age, and suggests that EHR exposure increases the likelihood of working in an office in the year after exposure by .015 percentage points, a 2.6\% increase relative to the mean. Splitting the sample by age reveals that this result is driven by the younger sample of physicians. There is no evidence that older physicians are shifting towards offices. This may be due to senior physicians already being established in some capacity in offices prior to EHR exposure, but is also supported by intuition that older workers are more "stuck" to their particular work environment.

If senior physicians are already established in offices in some capacity, they may be more affected on an intensive margin, the fraction of patients seen in an office relative to hospital or other settings. The estimates for this outcome are shown in Figure \ref{fig:officesecond}, where I find limited evidence that EHR exposure affects the fraction of patients seen in office settings. Of course, this could mean multiple things. If total number of patients remains constant, then this implies that the absolute number of patients seen in offices also is unchanged. However, if the total number of patients seen (which depends on efficiency in hospitals also) changes, then the absolute number of patients seen in offices necessarily changes in a similar direction. I revisit the implications of this result in Section \ref{sec:patientcount}, where I examine the total number of patients as an outcome variable.

The previous results only capture switching behavior which involves office-based settings. A more general way to examine physician switching behavior is to investigate whether the zip codes in which the physicians work is changing, which captures both office and other switching behaviors. Again, I exclude physicians who ever retire. The outcome variable is equal to 1 if the list of primary zip codes (up to twelve of them) changes. This variable is limited in the way it captures switching behavior because multiple scenarios are flagged as changing zip codes: leaving a particular zip code, adding one new zip code but leaving the rest unchanged, or completely changing zip codes. While this is a vague description of behavior, it provides a different insight into changes occurring because of EHR implementation.

@@

\subsection{Patient Count}\label{sec:patientcount}

An important, and heavily debated, aspect of the implementation of EHRs is whether or not they increase productivity in users. I measure productivity as the number of Medicare patients seen annually. EHRs could increase this number if they improve efficiency through less paperwork or ease of information access, or EHRs could decrease the number of patients seen if physicians are spending more time interfacing with the system and are not able to see as many patients.

In this case, productivity has two key determinants: the choice of physicians to learn and utilize the technology, and the realized quality/innovation of electronic health records themselves. Regarding the first determinant, I will assume that when a physician continues to work in an electronic record utilizing-hospital, they utilize the electronic health record system fully. That is, there are no physicians who stay actively working in a hospital but choose not to partake in the electronic health record used by the hospital. To understand the second determinant, the value added of electronic health records themselves, the sample of physicians is limited to those who have a positive number of patients in hospitals even after EHR exposure.

@@

EHRs have the potential to change claim incentives if the administrative costs of billing decrease or if EHRs encourage billing for more items than otherwise would have been recorded.

@@

@@

\begin{figure}
    \centering
    \caption{Hospital EHR Adoption Over Time}
    \includegraphics[scale=.6]{Objects/hosp_treat.png}
    \label{fig:hosp_treat}
\end{figure}
